---
title: "Project"
author: "Nene Diallo"
date: "2024-12-04"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

```{r}
load("/Users/ari3/Downloads/SCF2022 (1).RData")
View(data)

library(dplyr)
library(haven)
library(forcats)
summary(data$INCOME)

```


```{r}
##General Group Hypothesis

#We are going to explore how financial health varies between young adults (ages 18-30) and middle age/elderly individuals (ages 31-65) while measuring variables such as student loans, debt to income ratios, financial literacy, family structures, etc.

#Null Hypothesis: The financial health of young adults is not going to differ from the financial health of middle age/elderly individuals.

#Alternate Hypothesis: Middle age and elderly individuals will be more financially responsible and therefore they will have a better financial health than young adults.  

```

SETTING AGE SUBGROUPS
```{r}
# Subgroup: Young Adults (18-30)
young_adults <- data %>%
  filter(AGE >= 18 & AGE <= 30)

# Subgroup: Middle Age/Elderly (31-65)
middle_age_elderly <- data %>%
  filter(AGE >= 31 & AGE <= 65)

# Check the first few rows of each subset to ensure the division worked
head(young_adults)
head(middle_age_elderly)

View(young_adults)
View(middle_age_elderly)
```

RECODING SPENDMOR VARIABLE
```{r}
# Ensure SPENDMOR is a factor
data$SPENDMOR <- as.factor(data$SPENDMOR)

# Remove rows where SPENDMOR is "4" or "5"
data <- data %>%
  filter(!(SPENDMOR %in% c("4", "5")))

# Recode SPENDMOR to Spending_numeric
data$Spending_numeric <- forcats::fct_recode(data$SPENDMOR, 
                                             "Spending Exceeding Income" = "1", 
                                             "Spending Equal to Income" = "2", 
                                             "Spending Less than Income" = "3")

# Check the recoded column
summary(data$Spending_numeric)

# View the unique values to confirm correctness
table(data$Spending_numeric)

# Subgroup: Young Adults (18-30)
young_adults <- data %>%
  filter(AGE >= 18 & AGE <= 30) %>%
  mutate(Spending_numeric = forcats::fct_recode(SPENDMOR, 
                                                "Spending Exceeding Income" = "1", 
                                                "Spending Equal to Income" = "2", 
                                                "Spending Less than Income" = "3"))

# Subgroup: Middle Age/Elderly (31-65)
middle_age_elderly <- data %>%
  filter(AGE >= 31 & AGE <= 65) %>%
  mutate(Spending_numeric = forcats::fct_recode(SPENDMOR, 
                                                "Spending Exceeding Income" = "1", 
                                                "Spending Equal to Income" = "2", 
                                                "Spending Less than Income" = "3"))

# Check the results for each subgroup
summary(young_adults$Spending_numeric)
summary(middle_age_elderly$Spending_numeric)

```

RECODE FAMILY STRUCTURE
```{r}
#FAMSTRUCT recode

data$FAMSTRUCT <- as.factor(data$FAMSTRUCT)
data <- data %>% filter(FAMSTRUCT != "5")
data$Family_numeric <- fct_recode(data$FAMSTRUCT, 
                                "Single Individual" = "1", 
                                "Married" = "2", 
                                "Single Parent" = "3", 
                                "Other Family Type" = "4")
# Remove NA values from the Family_numeric column
data <- data %>% filter(!is.na(Family_numeric))

# View the resulting column of recoded values
summary(data$Family_numeric)
```


RECODE ASSET VARIABLE
```{r}
# Recode ASSET
data$asset_midpoint <- ifelse(data$ASSET == 0, "No assets",
                       ifelse(data$ASSET < 10000, "Less than $10k",
                       ifelse(data$ASSET >= 10000 & data$ASSET < 25000, "$10k - $24.9k",
                       ifelse(data$ASSET >= 25000 & data$ASSET < 50000, "$25k - $49.9k",
                       ifelse(data$ASSET >= 50000 & data$ASSET < 100000, "$50k - $99.9k",
                       ifelse(data$ASSET >= 100000 & data$ASSET < 500000, "$100k - $499.9k",
                       ifelse(data$ASSET >= 500000 & data$ASSET < 1000000, "$500k - $999.9k",
                       ifelse(data$ASSET >= 1000000 & data$ASSET < 5000000, "$1M - $4.9M",
                       ifelse(data$ASSET >= 5000000, "$5M+", NA)))))))))

# Handle NA values in asset_midpoint
data$asset_midpoint[is.na(data$asset_midpoint)] <- "Unknown"

# Subgroup: Young Adults
young_adults <- data %>%
  filter(AGE >= 18 & AGE <= 30)

# Check young_adults
str(young_adults)
summary(young_adults$asset_midpoint)

# Ensure asset_midpoint is ordered
young_adults$asset_midpoint <- factor(young_adults$asset_midpoint,
                                      levels = c("No assets", "Less than $10k", "$10k - $24.9k", "$25k - $49.9k",
                                                 "$50k - $99.9k", "$100k - $499.9k", "$500k - $999.9k",
                                                 "$1M - $4.9M", "$5M+", "Unknown"))


# Subgroup: Middle Age/Elderly
middle_age_elderly <- data %>%
  filter(AGE >= 31 & AGE <= 65)

# Check young_adults
str(middle_age_elderly)
summary(middle_age_elderly$asset_midpoint)

# Ensure asset_midpoint is ordered
middle_age_elderly$asset_midpoint <- factor(middle_age_elderly$asset_midpoint,
                                      levels = c("No assets", "Less than $10k", "$10k - $24.9k", "$25k - $49.9k",
                                                 "$50k - $99.9k", "$100k - $499.9k", "$500k - $999.9k",
                                                 "$1M - $4.9M", "$5M+", "Unknown"))
```

GRAPH FOR ASSETS VS SPENDING HABIT (YOUNG ADULTS)
```{r}
# Filter for Young Adults
young_adults_data <- summary_data %>% filter(subgroup == "Young Adults")

# Create the plot for Young Adults
p <- ggplot(young_adults_data, aes(x = asset_midpoint, y = Count, fill = Spending_numeric)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = custom_colors) +
  labs(
    title = "Comparison of Asset Levels and Spending Habits (Young Adults)",
    x = "Asset Levels",
    y = "Number of People",
    fill = "Spending Habits"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Convert to interactive plot
ggplotly(p)
```

GRAPH FOR ASSETS VS SPENDING HABIT (MIDDLE AGE/ELDERLY)
```{r}
# Load necessary libraries
library(dplyr)
library(ggplot2)
library(plotly)
library(forcats)

# Define custom colors for Spending_numeric levels
custom_colors <- c(
  "Spending Exceeded Income" = "pink",
  "Spending Equaled Income" = "purple",
  "Spending Less Than Income" = "blue"
)

# Filter for Middle Age/Elderly and clean the data
middle_age_elderly_data <- summary_data %>%
  filter(subgroup == "Middle Age/Elderly") %>% 
  mutate(
    # Ensure Spending_numeric matches the color keys
    Spending_numeric = forcats::fct_recode(
      Spending_numeric,
      "Spending Exceeded Income" = "1",
      "Spending Equaled Income" = "2",
      "Spending Less Than Income" = "3"
    )
  ) %>%
  filter(!is.na(Spending_numeric)) # Remove rows with NA in Spending_numeric

# Create the plot for Middle Age/Elderly
p <- ggplot(middle_age_elderly_data, aes(x = asset_midpoint, y = Count, fill = Spending_numeric)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = custom_colors) +  # Apply custom colors
  labs(
    title = "Comparison of Asset Levels and Spending Habits (Middle Age/Elderly)",
    x = "Asset Levels",
    y = "Number of People",
    fill = "Spending Habits"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Convert to interactive plot
ggplotly(p)
```


RECODE HOUSING COSTS (HOW MUCH EACH PERSON PAID FOR THEIR HOUSE)
```{r}
data$houses_midpoint <- ifelse(data$ASSET >= 0 & data$ASSET < 100000, "Less than $100k",
                       ifelse(data$ASSET >= 100000 & data$ASSET < 500000, "$100k - $500k",
                       ifelse(data$ASSET >= 500000 & data$ASSET < 1000000, "$500k - $1M",
                       ifelse(data$ASSET >= 1000000 & data$ASSET < 5000000, "$1M - $5M",
                       ifelse(data$ASSET >= 5000000 & data$ASSET < 10000000, "$5M - $10M",
                       ifelse(data$ASSET >= 10000000, "$5M+", NA))))))


data$houses_midpoint[is.na(data$houses_midpoint)] <- "Unknown"

# Subgroup: Young Adults
young_adults <- data %>%
  filter(AGE >= 18 & AGE <= 30)

# Check young_adults
str(young_adults)
summary(young_adults$houses_midpoint)


young_adults$houses_midpoint <- factor(young_adults$houses_midpoint,
                                      levels = c("Less than $100k", "$100k - $500k", "$500k - $1M", "$1M - $5M",
                                                 "$5M - $10M", "$10M+", "Unknown")) 


# Subgroup: Middle Age/Elderly
middle_age_elderly <- data %>%
  filter(AGE >= 31 & AGE <= 65)

# Check young_adults
str(middle_age_elderly)
summary(middle_age_elderly$houses_midpoint)

# Ensure asset_midpoint is ordered
middle_age_elderly$houses_midpoint <- factor(middle_age_elderly$houses_midpoint,
                                      levels = c("Less than $100k", "$100k - $500k", "$500k - $1M", "$1M - $5M",
                                                 "$5M - $10M", "$10M+", "Unknown")) 
```

GRAPH FOR HOUSING PRICE PAID (YOUNG ADULTS)
```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)

# Filter and recode for young adults (18-30)
young_adults <- data %>%
  filter(AGE >= 18 & AGE <= 30) %>%
  mutate(
    Family_numeric = factor(FAMSTRUCT, levels = c("1", "2", "3", "4"),
                            labels = c("Single Individual", "Married", "Single Parent", "Other Family Type")),
    houses_midpoint = cut(HOUSES,
                          breaks = c(-Inf, 100000, 500000, 1000000, 5000000, 10000000, Inf),
                          labels = c("Less than $100k", "$100k - $500k", "$500k - $1M",
                                     "$1M - $5M", "$5M - $10M", "$10M+"),
                          right = FALSE)
  )

# Plot for young adults
plot_young_adults <- ggplot(young_adults, aes(x = houses_midpoint, fill = Family_numeric)) +
  geom_bar(position = "dodge") +
  labs(
    title = "Housing Costs by Family Structure (Ages 18-30)",
    x = "How Much Did They Pay?",
    y = "Count",
    fill = "Household Family Structures"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Convert the ggplot object to an interactive plot
interactive_plot_young_adults <- ggplotly(plot_young_adults)

# Display the interactive plot
interactive_plot_young_adults

```


GRAPH FOR HOUSING PRICE PAID (MIDDLE AGE/ELDERLY)
```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)

# Filter and recode for middle-aged/elderly (31-65)
middle_age_elderly <- data %>%
  filter(AGE >= 31 & AGE <= 65) %>%
  mutate(
    Family_numeric = factor(FAMSTRUCT, levels = c("1", "2", "3", "4"),
                            labels = c("Single Individual", "Married", "Single Parent", "Other Family Type")),
    houses_midpoint = cut(HOUSES,
                          breaks = c(-Inf, 100000, 500000, 1000000, 5000000, 10000000, Inf),
                          labels = c("Less than $100k", "$100k - $500k", "$500k - $1M",
                                     "$1M - $5M", "$5M - $10M", "$10M+"),
                          right = FALSE)
  )

# Plot for middle-aged/elderly
plot_middle_age_elderly <- ggplot(middle_age_elderly, aes(x = houses_midpoint, fill = Family_numeric)) +
  geom_bar(position = "dodge") +
  labs(
    title = "Housing Costs by Family Structure (Ages 31-65)",
    x = "How Much Did They Pay?",
    y = "Count",
    fill = "Household Family Structures"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Convert the ggplot object to an interactive plot
interactive_plot_middle_age_elderly <- ggplotly(plot_middle_age_elderly)

# Display the interactive plot
interactive_plot_middle_age_elderly
```

COMBINING FOOD EXPENSES
```{r}
# Create a new column for total food expenses
data <- data %>%
  mutate(TotalFoodExpenses = FOODHOME + FOODAWAY + FOODDELV)

# Subgroup: Young Adults
young_adults <- data %>%
  filter(AGE >= 18 & AGE <= 30)

# Subgroup: Middle Age/Elderly
middle_age_elderly <- data %>%
  filter(AGE >= 31 & AGE <= 65)

# Recode Food Expense
data$food_expense <- ifelse(data$TotalFoodExpenses >= 0 & data$ASSET < 10000, "Less than $10k",
                       ifelse(data$ASSET >= 10000 & data$ASSET < 25000, "$10k - $25k",
                       ifelse(data$ASSET >= 25000 & data$ASSET < 50000, "$25k - $50k",
                       ifelse(data$ASSET >= 50000 & data$ASSET < 75000, "$50k - $75k",
                       ifelse(data$ASSET >= 75000 & data$ASSET < 100000, "$75k - $100k",
                       ifelse(data$ASSET >= 100000, "$100k+", NA))))))

# Handle NA values in asset_midpoint
data$food_expense[is.na(data$food_expense)] <- "Unknown"

# Subgroup: Young Adults
young_adults <- data %>%
  filter(AGE >= 18 & AGE <= 30)

# Check young_adults
str(young_adults)
summary(young_adults$food_expense)

# Ensure asset_midpoint is ordered
young_adults$food_expense <- factor(young_adults$food_expense,
                                      levels = c("Less than $10k", "$10k - $25k", "$25k - $50k", "$50k - $75k",
                                                 "$75k - $100k", "$100k+", "Unknown")) 


# Subgroup: Middle Age/Elderly
middle_age_elderly <- data %>%
  filter(AGE >= 31 & AGE <= 65)

# Check young_adults
str(middle_age_elderly)
summary(middle_age_elderly$food_expense)

# Ensure asset_midpoint is ordered
middle_age_elderly$houses_midpoint <- factor(middle_age_elderly$houses_midpoint,
                                      levels = c("Less than $10k", "$10k - $25k", "$25k - $50k", "$50k - $75k",
                                                 "$75k - $100k", "$100k+", "Unknown"))  
```

GRAPH FOR FOOD EXPENSES (YOUNG ADULTS) 
```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)

# Filter and recode for young adults (18-30)
young_adults <- data %>%
  filter(AGE >= 18 & AGE <= 30) %>%
  mutate(
    Family_numeric = factor(FAMSTRUCT, levels = c("1", "2", "3", "4"),
                            labels = c("Single Individual", "Married", "Single Parent", "Other Family Type")),
    food_expense = cut(TotalFoodExpenses,
                          breaks = c(-Inf, 10000, 25000, 50000, 75000, 100000, Inf),
                          labels = c("Less than $10k", "$10k - $25k", "$25k - $50k", "$50k - $75k",
                                                 "$75k - $100k", "$100k+"),
                          right = FALSE)
  )

# Plot for young adults
plot_young_adults <- ggplot(young_adults, aes(x = food_expense, fill = Family_numeric)) +
  geom_bar(position = "dodge") +
  labs(
    title = "Food Expenses by Family Structure (Ages 18-30)",
    x = "Food Expenses",
    y = "Count",
    fill = "Household Family Structures"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Convert the ggplot object to an interactive plot
interactive_plot_young_adults <- ggplotly(plot_young_adults)

# Display the interactive plot
interactive_plot_young_adults
```

GRAPH FOR FOOD EXPENSE (MIDDLE AGE/ELDERLY)
```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)

# Filter and recode for young adults (18-30)
middle_age_elderly <- data %>%
  filter(AGE >= 31 & AGE <= 65) %>%
  mutate(
    Family_numeric = factor(FAMSTRUCT, levels = c("1", "2", "3", "4"),
                            labels = c("Single Individual", "Married", "Single Parent", "Other Family Type")),
    food_expense = cut(TotalFoodExpenses,
                          breaks = c(-Inf, 10000, 25000, 50000, 75000, 100000, Inf),
                          labels = c("Less than $10k", "$10k - $25k", "$25k - $50k", "$50k - $75k",
                                                 "$75k - $100k", "$100k+"),
                          right = FALSE)
  )

# Plot for young adults
plot_middle_age_elderly <- ggplot(middle_age_elderly, aes(x = food_expense, fill = Family_numeric)) +
  geom_bar(position = "dodge") +
  labs(
    title = "Food Expenses by Family Structure (Ages 31-65)",
    x = "Food Expenses",
    y = "Count",
    fill = "Household Family Structures"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Convert the ggplot object to an interactive plot
interactive_plot_middle_age_elderly <- ggplotly(plot_middle_age_elderly)

# Display the interactive plot
interactive_plot_middle_age_elderly
```

LOGIT MODEL
```{r}
library(dplyr)

# Recode Family_numeric for the entire dataset
data <- data %>%
  mutate(Family_numeric = ifelse(Family_numeric == "Single", 1, 0)) # Adjust as needed

# Split data into subgroups
young_adults <- data %>% filter(AGE >= 18 & AGE <= 30)
middle_age_elderly <- data %>% filter(AGE >= 31 & AGE <= 65)

# Logistic regression for young adults
young_adults_logit <- glm(Family_numeric ~ asset_midpoint + houses_midpoint + food_expense, 
                          data = young_adults, 
                          family = binomial(link = "logit"))

# Logistic regression for middle-age/elderly
middle_age_elderly_logit <- glm(Family_numeric ~ asset_midpoint + houses_midpoint + food_expense, 
                                data = middle_age_elderly, 
                                family = binomial(link = "logit"))

# Summarize the models
summary(young_adults_logit)
summary(middle_age_elderly_logit)

```

TYPE 1 TYPE 2 ERROR - YOUNG ADULTS
```{r}
pred_vals <- predict(young_adults_logit, young_adults, type = "response")
pred_young_adults_logit <- (pred_vals > mean(pred_vals))
table(pred = pred_young_adults_logit, true = young_adults$Family_numeric)
```
TYPE 1 TYPE 2 ERROR - MIDDLE AGE/ELDERLY
```{r}
pred_vals <- predict(middle_age_elderly_logit, middle_age_elderly, type = "response")
pred_middle_age_elderly_logit <- (pred_vals > mean(pred_vals))
table(pred = pred_middle_age_elderly_logit, true = middle_age_elderly$Family_numeric)
```

INCOME RECODE
```{r}
#Income recode
data$income_midpoint <- ifelse(data$INCOME < 25000, "HH income less than $25k",
                        ifelse(data$INCOME >= 25000 & data$INCOME < 35000, "HH income $25k - $34.9k",
                        ifelse(data$INCOME >= 35000 & data$INCOME < 50000, "HH income $35k - $49.9k",
                        ifelse(data$INCOME >= 50000 & data$INCOME < 75000, "HH income $50k - $74.9k",
                        ifelse(data$INCOME >= 75000 & data$INCOME < 100000, "HH income $75k - $99.9k",
                        ifelse(data$INCOME >= 100000 & data$INCOME < 150000, "HH income $100k - $149.9k",
                        ifelse(data$INCOME >= 150000 & data$INCOME < 200000, "HH income $150k - $199.9k",
                        ifelse(data$INCOME >= 200000 & data$INCOME < 500000, "HH income $200k - $499.9k",
                        ifelse(data$INCOME >= 500000 & data$INCOME < 1000000, "HH income $500k - $999.9k", 
                               "HH income $1M+")))))))))

# Subgroup: Young Adults
young_adults <- data %>%
  filter(AGE >= 18 & AGE <= 30)

# Check young_adults
str(young_adults)
summary(young_adults$income_midpoint)

# Ensure asset_midpoint is ordered
young_adults$income_midpoint <- factor(young_adults$income_midpoint,
                                      levels = c("HH income $25k - $34.9k", "HH income $35k - $49.9k", "HH income $50k - $74.9k" ,"HH income $75k - $99.9k","HH income $100k - $149.9k", "HH income $150k - $199.9k", "HH income $200k - $499.9k", "HH income $500k - $999.9k")) 


# Subgroup: Middle Age/Elderly
middle_age_elderly <- data %>%
  filter(AGE >= 31 & AGE <= 65)

# Check young_adults
str(middle_age_elderly)
summary(middle_age_elderly$income_midpoint)

# Ensure asset_midpoint is ordered
middle_age_elderly$income_midpoint <- factor(middle_age_elderly$income_midpoint,
                                      levels = c("HH income $25k - $34.9k", "HH income $35k - $49.9k", "HH income $50k - $74.9k" ,"HH income $75k - $99.9k","HH income $100k - $149.9k", "HH income $150k - $199.9k", "HH income $200k - $499.9k", "HH income $500k - $999.9k"))  
```

FOOD EXPENSE VS INCOME
```{r}
library(ggplot2)

# Create scatter plot with limited ranges for income and food expenses
ggplot(young_adults, aes(x = INCOME, y = TotalFoodExpenses)) +
  geom_point(alpha = 0.6, color = "pink") +
  geom_smooth(method = "lm", se = TRUE, color = "darkgreen") + # Optional trend line
  labs(
    title = "Scatter Plot of Food Expenses vs. Income (Young Adults)",
    x = "Income",
    y = "Food Expenses Yearly"
  ) +
  scale_x_continuous(limits = c(0, 100000), labels = scales::comma) + # Restrict x-axis range
  scale_y_continuous(limits = c(0, 50000), labels = scales::comma) + # Restrict y-axis range
  theme_minimal()



# Create scatter plot with limited ranges for income and food expenses
ggplot(middle_age_elderly, aes(x = INCOME, y = TotalFoodExpenses)) +
  geom_point(alpha = 0.6, color = "pink") +
  geom_smooth(method = "lm", se = TRUE, color = "darkgreen") + # Optional trend line
  labs(
    title = "Scatter Plot of Food Expenses vs. Income (Middle Age/Elderly)",
    x = "Income",
    y = "Food Expenses Yearly"
  ) +
  scale_x_continuous(limits = c(0, 100000), labels = scales::comma) + # Restrict x-axis range
  scale_y_continuous(limits = c(0, 50000), labels = scales::comma) + # Restrict y-axis range
  theme_minimal()
```


```{r}
library(ggplot2)

# Create scatter plot with limited ranges for income and food expenses
ggplot(young_adults, aes(x = INCOME, y = RENT)) +
  geom_point(alpha = 0.6, color = "violet") +
  geom_smooth(method = "lm", se = TRUE, color = "orange") + # Optional trend line
  labs(
    title = "Scatter Plot of Rent vs. Income (Young Adults)",
    x = "Income",
    y = "Monthly Rent"
  ) +
  scale_x_continuous(limits = c(0, 100000), labels = scales::comma) + # Restrict x-axis range
  scale_y_continuous(breaks = seq(0, 3000, by = 500), labels = scales::comma) + # Restrict y-axis range
  theme_minimal()




library(scales)

ggplot(middle_age_elderly, aes(x = INCOME, y = RENT)) +
  geom_point(alpha = 0.6, color = "violet") +
  geom_smooth(method = "lm", se = TRUE, color = "orange") + # Optional trend line
  labs(
    title = "Scatter Plot of Rent vs. Income (Middle Age/Elderly",
    x = "Income",
    y = "Monthly Rent"
  ) +
  scale_x_continuous(limits = c(0, 100000), labels = comma) + # Restrict x-axis range
  scale_y_continuous(limits = c(0, 5000, by = 500), labels = comma) + # Automatic breaks
  theme_minimal()


```

